\subsection*{Preface}
\addcontentsline{toc}{subsection}{Acknowledgments} 

The initial motivation for our work with MARSS models was a collaboration with Rich Hinrichsen. Rich developed a framework for analysis of multi-site population count data using MARSS models and bootstrap AICb \citep{HinrichsenHolmes2009}. Our work (EEH and EJW) extended Rich's framework, made it more general, and led to the development of a parametric bootstrap AICb\index{model selection!bootstrap AIC, AICbp} for MARSS models, which allows one to do model-selection using datasets with missing values \citep{Wardetal2009c, HolmesWard2010}.  Later, we developed additional algorithms for simulation and confidence intervals.  Discussions with Mark Scheuerell led to an extensive revision of the EM algorithm and to the development of a general EM algorithm for constrained MARSS models \citep{Holmes2010}.  Discussions with Mark also led to a complete rewrite of the model specification so that the package could be used for MARSS models in general---rather than simply the form of MARSS model used in our (EEH and EJW) applications.  Many collaborators have helped test the package; we thank especially Yasmin Lucero, Mark Scheuerell, Kevin See, and Brice Semmens.  Development of the code into a R package would not have been possible without Kellie Wills, who wrote the much of the code outside of the algorithm functions.

The case studies used in this manual were developed for workshops on analysis of multivariate time-series data given at the Ecological Society meetings since 2005 and taught by us (EEH and EJW) along with Yasmin Lucero, Stephanie Hampton, Brice Semmens, and Mark Scheuerell.  The case study on extinction estimation and trend estimation was initially developed by Brice Semmens and later extended by us for this manual. The algorithm behind the TMU figure in Chapter \ref{chap:CSpva} was developed during a collaboration with Steve Ellner \citep{EllnerHolmes2008}.
 
EEH and EJW are research scientists at the Northwest Fisheries Science Center in the Mathematical Biology program.  This work was conducted as part of our jobs at the Northwest Fisheries Science Center, a research center for NOAA Fisheries which is a US federal government agency.   A CAMEO grant from NOAA Fisheries supported Kellie Wills.  During the initial stages of this work, EJW was supported on a post-doctoral fellowship from the National Research Council.

You are welcome to use the code and adapt it with attribution.  It may not be used in any commercial applications.  
Links to more code and publications on MARSS applications can be found by following the links at EEH's website http://faculty.washington.edu/eeholmes.   Links to our papers that use these methods can also be found at the same website.